---
title: "dplyr"
author: "bhaskar"
output: html_document
---

```{r}
suppressMessages(library(dplyr))
library(hflights)
data(hflights)
head(hflights)
```

**tbl_df creates "local data frame"**
```{r}
flights = tbl_df(hflights)
flights
# specify number of rows
print(flights,n = 20)
# convert to a normal data frame to see all the columns
data.frame(head(flights))
```

## filter: Keep rows of matching criteria

for all dplyr verbs first argument and return type is data frame
```{r}
filter(flights,Month==1,DayofMonth==1)
filter(flights,UniqueCarrier=="AA"|UniqueCarrier == "AS")
filter(flights,UniqueCarrier %in% c("AA","AS"))
```

## select: pick columns by name

```{r}
select(flights,DepTime,ArrTime,FlightNum)
# use colons for multiple continous columns ans use contains to match
# starts_with, ends_with and regEx + matches can be used for matching
select(flights,Year:DayofMonth,contains("Taxi"),contains("Delay"))
select(flights,Year:DayofMonth,contains("Taxi"),contains("delay"))
```

**Chaining or pipelining is done using %>% operator**
```{r}
flights %>% select(UniqueCarrier,DepDelay) %>% filter(DepDelay>60)
x1 = 1:5; x2 = 2:6 ; 
sqrt(sum((x1-x2)^2))
(x1-x2)^2 %>% sum() %>% sqrt()
# flights %>% select(flights,UniqueCarrier,DepDelay) %>%  filter(flights,DepDelay>60)
```

## arrange: reorder Rows
```{r}
flights %>% select(UniqueCarrier,DepDelay) %>% arrange(DepDelay)
flights %>% select(UniqueCarrier,DepDelay) %>% arrange(desc(DepDelay))
```

## mutate : Add new variables
```{r}
flights = flights %>% mutate(Speed = Distance/AirTime*60)
```
## summarise : Reduce variables to values
primarily useful with the data that has been grouped by one or more variables
group_by creates the group that will be operated
summarise uses the provided aggregation
```{r}
flights %>% group_by(Dest) %>% summarise(avg_delay = mean(ArrDelay,na.rm = T))
```
**summarise_each** allows to apply same summary function to multiple columns at once
```{r}
flights %>% group_by(UniqueCarrier) %>% summarise_each(funs(mean),Cancelled,Diverted)
flights %>%  group_by(UniqueCarrier) %>% summarise_each(funs(min(.,na.rm=T),max(.,na.rm=t)),matches("Delay"))
```
**n() and n_distinct()**
```{r}
flights %>% group_by(Month,DayofMonth) %>% summarise(flight_count = n_distinct(Dest)) %>% arrange(desc(flight_count))
flights %>% group_by(Month,DayofMonth) %>% summarise(flight_count = n()) %>% arrange(desc(flight_count))
flights %>% group_by(Dest) %>% summarise(flights = n(),distinct = n_distinct(TailNum))
```
**tally()** function
```{r}
flights %>% group_by(Month,DayofMonth) %>% tally(sort = T)
```

*sometimes grouping is useful without summarising*
```{r}
flights%>%group_by(Dest)%>%select(Cancelled)%>%table()%>%head()
```

### Window functions
Five main types of windows functions, 3 useful for R
1. Ranking functions like *row_number(),min_rank(),dense_rank(),cume_dist(),percent_rank()
2. Offset functions lead and lag
3. variation of aggregate functions like cumsum(),cummin(),cummax(),cumall(),cummany()
the other two types fixed width aggregation and recycled aggregation are less useful.
```{r}
# for each carrier calculate the two days of the year they had their longest departure delays
flights %>% group_by(UniqueCarrier) %>% select(DepDelay,DayofMonth,Month) %>% filter(min_rank(desc(DepDelay))<=2) %>% arrange(DepDelay,UniqueCarrier)
# top_n() uses filter and min_rank together, variable used for ordering is by default the last column
flights %>% group_by(UniqueCarrier) %>% select(DayofMonth,Month,DepDelay) %>% top_n(2) %>% arrange(desc(DepDelay),UniqueCarrier)
flights %>% group_by(UniqueCarrier) %>% select(DayofMonth,Month,DepDelay) %>% top_n(2,Month) %>% arrange(desc(DepDelay),UniqueCarrier)
# for each month, calculate the number of flights and the change from the previous month
flights %>% group_by(Month) %>% summarise(flights_num = n()) %>% mutate(change=flights_num - lag(flights_num))
#using tally, remember tally gives count of each group in column 'n'
flights %>% group_by(Month) %>% tally() %>% mutate(change=n-lag(n))
```

*other useful functions*
```{r}
# randomly sample a fixed number of rows, without replacement
flights %>% sample_n(5)
 #randomly sample a fraction of rows, with replacement
flights %>% sample_frac(0.25, replace=TRUE)
head(flights)
glimpse(flights)
inner_join(flights,flights)
# semi_join90, anti_join(),
```

## tidyr functions
```{r}
library(tidyr)
flights %>% select(UniqueCarrier,TailNum)%>%mutate(TailLink=paste(TailNum,sample(1:4,n(),replace = T))) %>% separate(TailLink,c("a","b")) %>% head()
```
seperate() X unite to seperate or unite multiple columns based on splitting single columns
gather() aX spread() to make wide data longer or viceversa
